Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-03 response_to_reviewers_R1.Rmd

---
title: "Response to Reviewers"
date: "2023-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

## Major comments from Reviewer 2

Overall Summary:

This revised version is a significant improvement over the previous version and the reviewers have addressed my major concerns with the paper with one exception.

Suggested Changes:

The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.

I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place. This could be moved to Section 7 (Opportunities). Likewise, Section 5 could be retitled something like‘Betting Data and Sports Analytics” or “Betting Data Analysis”.

## Minor Changes from Reviewer 2

- Page 1, line 47 “And, of course, in baseball” add commas before and - after “of course”

> Done.

- Page 2, line 5, “games share” instead of “game shares”

> Fixed.

- Page 2 line 6, “anthropologically’ seems odd here. Maybe “historically” instead

> Fixed.

- Page 8, Macdonald (2012?) used expected possession value for ice hockey to evaluate players.

> We have a added a citation to this work, but note that whereas Macdonald is using expected *goals*, he did not have access to player tracking data necessary to compute an EPV metric.

- Section 6.1, line 53, Figure ??
- Section 6.1, line 56, contextual is used twice in that sentence, perhaps “helping the reader visualize” or “helping the reader understand”

> We have changed this to "understand".

- Table 3, the third column of this table is off by a line with the entries in the first two columns.

> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.

- Section 6.2, line 52, Figure 6 should be Figure 7; likewise the figure number in the next sentence

> This is a downstream consequence of the previous error with Figure ??

- Figure 6, in my version the images are stacked rather than left/right
- Section 6.2, line 46 This sentence that starts “Consider, for example, “ seems out of place with the rest of the tone of this article
- Section 7, line 7 or 8, it might be worth mentioning the growth of small conferences like UCSAS that have sprung up, something like “As the field of sports analytics has grown, there has been a proliferation of sports specific and regional sports analytics conferences.”

> We have added a similar sentence to this effect.

### 2023-01-03 sports_analytics_R0.Rmd
start on #44

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources. 
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
link-citations: yes
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
```

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR",
    "sportyR",
    "nflplotR",
    "baseballr"
  ),
  file = "pkgs.bib"
)

@@

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].

@@

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, we discuss the use of non-performance data (most notably, sports market betting data) in Section \ref{sec:betting}, and in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

The concept of a state is easier to grasp in a sport that can be modeled as *discrete* (in the sense of [discrete event simulation](https://en.wikipedia.org/wiki/Discrete-event_simulation)). 
By discrete, we mean a sport that can be easily broken into short, distinct segments of action which are typically summarized categorically. 
Each of these segments might represent a state $s$. 
For example, each pitch in baseball is either a ball or a strike. 
If the ball is put in play, then there may be a complex sequence of movements by the players, but ultimately (within a few seconds) that sequence will end and no more action will be permitted until the next pitch. 
At the beginning and end of each phase of action, we will know definitively which team is on offense and defense, which runners are on which bases, the score, how many outs there are, etc. 
Tennis could similarly be viewed as a series of discrete actions defined by each point. 
To say that a player is winning 6-2, 3-1, 40-15 and serving with one fault committed is to characterize the state of the match.
In American football, the game can be broken down into a discrete sequence based on each down. 
Contrast this to sports like lacrosse, soccer, or any variant of hockey, which feature largely running clocks and continuous player movement. 
In these sports, it is not obvious how to break up the action into discrete chunks.

In this Section, we illustrate how the fundamental concept of the expected value of a game state leads to compelling findings in a variety of sports.

## Discrete event analysis

First, we explore results derived from the expected value of a state in sports where discrete event analysis is common. 
We draw primarily on baseball and American football, but applications in other sports (e.g., tennis) are common (see for example, @kovalchik2019calibration).

### In baseball, the expected run matrix

In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics.

@@

Early work on this topic can be found in @lindsey1963investigation, who used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning. 
This line of work led to analysis of all types of common baseball strategies. 
For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out. Figure \ref{fig:lindsey} shows a reproduction of @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its most common form.

@Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

### In American football, expected points {#sec:football}

@@

The task of estimating expected points of possession in football goes back to @Carter1971operations, who estimate the expected points for 1st and 10 plays in the NFL, given any yard line on the football field. 
Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
Figure \ref{fig:carter} shows a reproduction of @Carter1971operations's estimates. 
As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.
One limitation of this approach is the linearity assumption, which results in a high negative expected point value when the offensive team is 95 yards away from the opponent's goal line.

@@

Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other play outcomes. 
For any given play, the expected points are calculated using the absorption probabilities for different scoring events.

@@

### In American football, 4th down strategy

@@

Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a causal inference framework to reach similar conclusions that NFL teams are too conservative in going for it on 4th down.  In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "got for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e. 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1. 
In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM]. However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making. 
With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data. 
Many past analyses of the decision to go for it or not on 4th down conclude that coached in the NFL are too conservative in their decisions making. 
@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but the effect may be overstated due to the way in which to go yardage has been recorded in the past.

@@

## Continuous event analysis

@@

### In basketball, expected point value

@@

Note that Miami starts the play with an EPV of approximately 1.0 points, which indicates their implied average points per possession. 
Chalmers' shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court.

@@

See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
The strategy of maximizing expected points in basketball has led directly to the proliferation of three-point shooting in the NBA.

### In American football, yards gained

@@

### In other sports

@@

@kumagai2021hockey offer an EPV metric for ice hockey.

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u \in U$ that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Information derived from Table \ref{tab:erm} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the *defensive* team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

```{r, eval=FALSE}
erm <- erm |>
  mutate(
    Pr_0 = c(
      .747, .855, .933,
      .604, .734, .886, 
      .381, .610, .788,
      .120, .307, .738,
      .395, .571, .791,
      .130, .367, .717,
      .180, .270, .668,
      .180, .303, .671
    )
  )

@@

In addition, @Yurko2019nflwar estimate win probability in the NFL using a generalized additive model (GAM), as part of the **nflscrapR** package [@R-nflscrapR] and nflWAR framework. 
This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
The two win probability frameworks proposed by @Lock2014 and @Yurko2019nflwar were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in American football.

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(linewidth = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme(legend.position = "top")

@@

On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
@Deshpande2016 propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
This model assumes independence of observations and constant variability in win probability.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
The methods in this Section apply equally well to multiplayer teams and single-player teams.

The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.

## Expected winning percentage {#sec:wpct}

A simple method for estimating team strength that has become popular in sports analytics is expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.

Expected winning percentage is just:
$$
  \widehat{wpct} = \frac{X^\alpha}{X^\alpha + Y^\alpha} \,,
$$
where $X$ is the number of points (runs) that a team has scored, and $Y$ is the number of points (runs) that they have allowed, over some specified time period. 
James's work was originally in baseball, and he posited the value of $\alpha = 2$. 
The resemblance to the formula for computing the length of the hypotenuse in a right triangle provides the nod to Pythagoras.

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954, where the optimal value of $\alpha$ is 1.84.

```{r}
library(tidyverse)
library(Lahman)
exp_wpct <- function(x, k = 2) { 
  return(1/(1 + (1/x)^k))
}

TeamRuns <- Teams %>% 
  filter(yearID >= 1954) %>%
  rename(RS = R) %>% 
  mutate(WPct = W / (W + L), run_ratio = RS/RA) %>%
  select(yearID, teamID, lgID, WPct, run_ratio)

data_space <- ggplot(
  data = TeamRuns, aes(x = run_ratio, y = WPct)
) +
  geom_vline(xintercept = 1, color = "darkgray", linetype = 2) +
  geom_hline(yintercept = 0.5, color = "darkgray", linetype = 2) +
  geom_point(alpha = 0.2) + 
  xlab("Ratio of Runs Scored to Runs Allowed") + 
  ylab("Winning Percentage")
```

```{r}
optimal <- TeamRuns %>%
  nls(
    formula = WPct ~ 1/(1 + (1/run_ratio)^k), 
    start = list(k = 2)
  ) %>%
  coef()
```

```{r pythag, fig.cap="Winning percentages vs. runs scored and runs allowed in baseball, 1954--2021. The navy line represents the expected winning percentage model posited by James, with the exponent value of 2. The gold line shows the same model with an optimal exponent of 1.84."}
data_space +
  stat_function(fun = exp_wpct, args = list(2), linewidth = 2, color = "navy") +
  stat_function(fun = exp_wpct, args = list(optimal), linewidth = 2, color = "gold")
```

Many authors have fit expected winning percentage models to other sports---too many to cite here. 
See, for example, @hamilton2011extension for association football (soccer), @caro2013testing for Division I college football, and notably, future NBA general manager Daryl Morey for basketball [@dewan1993stats].

## Bradley-Terry models

Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{(\beta_i)}$ is often viewed as a representation of team $i$'s ability.

@@

tennis_bt |>
  update(refcat = "Wawrinka") |> 
  BradleyTerry2::BTabilities() |> 
  as_tibble(rownames = "Player") |>
  arrange(desc(ability)) |>
  mutate(
    Ability = as.double(ability),
    SE = as.double(s.e.)
  ) |>
  select(Player, Ability, SE) |>
  kable(
    booktabs = TRUE, 
    digits = 3,
    caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}"
  ) |> 
  kable_styling(latex_options = "hold_position")
```

@@

## Elo ratings {#sec:elo}

<!-- a basic Elo probability model for estimating the probability of $A$ beating $B$ is -->
<!-- $$ -->
<!-- P(A) = W(R_A - R_B) \,. -->
<!-- $$ -->
<!-- Here, $W$ is a continuous and monotonically increasing function that maps the difference in strength to (0, 1).  -->

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess.
Given two players $i$ and $j$ with unknown ratings $R_i$ and $R_j$, the probability $\Pi_{ij}$ of $i$ beating $j$ is defined as
$$
\Pi_{ij} = \frac{1}{1 + K^{(R_j - R_i)/400}} \,.
$$
In this formula, $K$ is commonly known as the $K$-factor, or development coefficient.
The International Chess Federation (FIDE) uses $K=10$ for players with any previously achieved rating of at least 2400.
For those that have never reached a rating of 2400 or better, $K=20$ is assigned. 
Finally, $K=40$ is given to new players with under 30 games played, and players under the age of 18 with rating fewer than 2300 [@fide2022rating].

Another interpretation for $\Pi_{ij}$ is the expected score of the game for player $i$.
The scores of 0, 0.5, and 1 are associated with three possible game outcomes loss, tie, and win, respectively.
After a game, the updated Elo rating $R^*_i$ for player $i$ is
$$
R^*_i = R_i + K(S_i - \Pi_{ij}) \,,
$$
where $S_i \in \{0, 0.5, 1\}$.
When a tournament concludes, a post-tournament rating is obtained for each player based on the rating updates for all games played.

```{r}
library(chessR)
us <- c("fatgreggy")
greg <- get_raw_chessdotcom(
  usernames = us,
  year_month = 202206
)
# greg: black
```

To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR] (see Section \ref{sec:chess}). 
Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas his opponent was rated `r greg[6, "WhiteElo"]`.
Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
The probability of the author ($a$) defeating their opponent ($b$) was
<!-- don't know how to include `r` inside latex -->
$$
\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
$$
The author won the match: that outcome is associated with a score of $S_a = 1$.
The post-game Elo rating for the author is thus
$$
R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
$$

Besides chess, the Elo system has also been implemented to estimate team strength in other sports. 
See @koning2017rating for more information on applications of the Elo rating in soccer, and @kovalchik2019calibration and @kovalchik2016searching for Elo ratings in tennis. 
Furthermore, Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).

## Bayesian state-space models

@Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources of variation in team strengths. 
See also @glickman2017estimating for more discussion on estimating team strengths in American football.

More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
Both papers motivate the usefulness of model-based measures of team strength by demonstrating their superiority to low-resolution win-loss records. 
Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect three big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources. 
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: 
  rticles::arxiv_article:
    keep_md: yes
header-includes:
  - \usepackage{amsmath}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
library(tidyverse)
library(kableExtra)
```

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR"
  ),
  file = "pkgs.bib"
)

@@

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact.

@@

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible.

@@

## In baseball, the expected run matrix

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of George @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

@@

Early work on this topic can be found in @lindsey1963investigation, who used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning. 
This line of work led to analysis of all types of common baseball strategies. 
For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out. 
@Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

## In American football {#sec:football}

@@

The task of estimating expected points of possession in football goes back to @Carter1971operations, who estimate the expected points for 1st and 10 plays in the NFL, given any yard line on the football field. 
Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
Figure \ref{fig:carter} shows a reproduction of @Carter1971operations's estimates. 
As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.

@@

Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
For any given play, the expected points are calculated using the absorption probabilities for different scoring events.

@@

### 4th down strategy

@@

Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.  In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.  Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go. They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.

@@

## In sports with continuous states

@@

@@

Note that Miami starts the play with an EPV of approximately 1.0 points, which indicates their implied average points per possession. 
Chalmers' shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
A breakthrough in this work is that this estimate is conditional on the locations of the other 9 players on the basketball court!

@@

See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
The strategy of maximizing expected points in basketball has lead directly to the proliferation of three-point shooting in the NBA.

@@

@@

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u \in U$ that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Table \ref{tab:tied} (derived from Table \ref{tab:erm}) reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the *defensive* team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

```{r}
erm <- erm |>
  mutate(
    Pr_0 = c(
      .747, .855, .933,
      .604, .734, .886, 
      .381, .610, .788,
      .120, .307, .738,
      .395, .571, .791,
      .130, .367, .717,
      .180, .270, .668,
      .180, .303, .671
    )
  )

@@

In addition, @Yurko2019nflwar estimate win probability in the NFL using a generalized additive model (GAM), as part of the **nflscrapR** package [@R-nflscrapR] and nflWAR framework. 
This model takes into account the estimated expected points obtained from the model described in Section \ref{sec:ev}, along with other predictors for time, current half, and timeouts. 
The two win probability frameworks proposed by @Lock2014 and @Yurko2019nflwar were also implemented in @Yam2019 with minimal modifications. Specifically, the authors combined both approaches to estimate the win probability for each play, with an overall goal of assessing fourth down decision-making in gridiron football.

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")

@@

On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
This model assumes independence of observations and constant variability in win probability.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations. 
Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.

@@

tennis_bt |>
  update(refcat = "Wawrinka") |> 
  BradleyTerry2::BTabilities() |> 
  round(3) |> 
  kable(booktabs = TRUE, 
        col.names = c("Ability", "SE"),
        caption = "The estimated abilities (with standard errors) for each tennis player, relative to Wawrinka, obtained from the fitted Bradley-Terry model. \\label{tab:tennisbtm}") |> 
  kable_styling(latex_options = "hold_position")
```

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
See @koning2017rating for more information on applications of the Elo rating in soccer.

@Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources of variation in team strengths. 
See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
Both papers motivate the usefulness of model-based measures of team strength by demonstrating their superiority to low-resolution win-loss records. 
Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.
